---
title: "From Points to Paychecks: An Exploratory Analysis of NBA Salaries and Performance (2022–23)"
author: " David Lin"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    code_folding: show
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{subcaption}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(tidyverse)
library(skimr)
library(GGally)
library(kableExtra)
library(janitor)
library(stargazer)
library(renv)
library(car)
library(lmtest)
library(sandwich)
library(dplyr)
library(knitr)
library(kableExtra)
library(tibble)
library(patchwork)
```

\newpage
# Introduction

GMs and coaches of NBA teams need to make highstakes decisions when assembling their rosters. With the average NBA player salary topping $6.62 million for the 2022-23 season ([1]), the need for data insights is higher than ever. Our goal is to describe how NBA player salaries relate to performance metrics during the 2022-23 **regular** season. We focus on Points Per Game (PTS) and Minutes Played (MP) as primary predictors, along with other stats like FG, TRB, 3P%, TOV, WS, AST, and GP. This exploratory analysis will derive actionable insights and enable key stakeholders to make data-driven decisions.

## Research Approach

While prior studies use machine learning for salary prediction ([2, 3]), our Ordinary Least Squares (OLS) regression model prioritizes interpretability 
to identify which performance indicators are most descriptively associated (not causal) with compensation. Because this is a team sport, we acknowledge 
the presence of dependence among observations and violations of the **IID** (independent and identically distributed) assumption. However, we proceed with the analysis given our focus on descriptive insights. We will revisit these hypotheses in our conclusion to confirm or refine their validity.

# Data Source

- **Primary Source:** [Kaggle NBA Salaries 2022-23](https://www.kaggle.com/datasets/jamiewelsh2/nba-player-salaries-2022-23-season/data)
- **All-Star Players:** [Basketball Reference All-Star 2023](https://www.basketball-reference.com/allstar/NBA_2023.html)
- **Years of Experience:** [Basketball Reference Team Rosters](https://www.basketball-reference.com/teams/ATL/2023.html)
- **Market Size Data:** [NBA Team Market Size Rankings](https://hoop-social.com/nba-team-market-size-rankings/)
- **Unit:** Individual NBA players
- **Type:** Cross-sectional snapshot of 2022-23 season

# Data Wrangling

For data wrangling, we began by obtaining the main dataset from Kaggle and iteratively scraping roster data for each team from Basketball Reference. We then added All-Star selections and merged the dataset with market size information. Duplicate entries, primarily from players traded mid-season, were identified and removed. The cleaned data was split into two parts: 30/70 for exploratory data analysis (EDA) and the remainder for confirmatory analysis. The current analysis focuses on the EDA portion.

```{r data-clean, echo=FALSE}
# Load and clean the data
nba_raw <- read_csv("../../data_nba/eda.csv")

# Clean and prepare the data
nba <- nba_raw %>%
  filter(!is.na(Salary), GP > 0) %>%
  select(Salary, PTS, MP, FG, TRB, `3P%`, TOV, WS, GP, AST, tv_market_size, `TS%`, 
         star_player, Position, Age, Exp, Team) %>%
  mutate(
    LogSalary = log(Salary),
    TVMS = tv_market_size,
    IsRookie = ifelse(Exp == 'R', 1, 0),
    Exp_num = as.numeric(replace(Exp, Exp == 'R', 0)),
    RookieOrStar = factor(case_when(
      Exp_num == 0 ~ 1, # Use Exp_num here
      star_player == 1 ~ 3,
      TRUE ~ 2
    ), levels = 1:3, labels = c("Rookie", "Regular", "Star"), ordered = TRUE),
    player_category = factor(case_when( # Define player_category here for consistency
      star_player == 1 ~ "Star",
      IsRookie == 1 ~ "Rookie",
      TRUE ~ "Regular"
    ),
    # levels = c("Rookie", "Regular", "Star")
    )
  )

# Fix data types and create Box-Cox transformed Salary
nba$star_player <- factor(nba$star_player)
# Exp is already numeric via Exp_num, no need to convert nba$Exp directly if Exp_num is used consistently
nba$Position <- factor(nba$Position)

lambda_obj <- powerTransform(nba$Salary)
nba <- nba %>% mutate(BoxCoxSalary = bcPower(Salary, lambda_obj$lambda))
```

The raw NBA dataset contains `r nrow(nba_raw)` players and `r ncol(nba_raw)` variables for the 2022-23 season. After cleaning, the dataset contains `r nrow(nba)` active players with valid salary data. We selected `r ncol(nba)` key variables including performance metrics, player characteristics, and team information. We understand that Salary is heavily right-skewed, so we created a log-transformed variable (LogSalary) to normalize the distribution. We also created a Box-Cox transformed variable (BoxCoxSalary) with lambda of `r lambda_obj$lambda` to further improve normality.

```{r check-nas, echo=FALSE}
# # Check for any remaining NAs
# na_summary <- colSums(is.na(nba))
# if(sum(na_summary) > 0) {
#   cat("Variables with missing values after cleaning:\n")
#   print(na_summary[na_summary > 0])
# } else {
#   cat("No missing values remain in the cleaned dataset.\n")
# }
```

# Operationalization

- **Salary:** Player compensation in USD
- **LogSalary:** Natural logarithm of salary (created to address right-skewness)
- **PTS:** Points per game = Total points / games played
- **MP:** Minutes played per game
- **Other metrics:** FG (field goals), TRB (rebounds per game), `3P%` (3-point shooting percentage), TOV (turnovers), WS (win shares), AST (assists per game), GP (games played)
- **TVMS:** TV Market Size - represents the size of the team's market
- **TS%:** True Shooting Percentage - a measure of shooting efficiency

# Exploratory Data Analysis

## Salary Distributions and Star Player Comparison

The salary distribution is heavily right-skewed (Figure \@ref(fig:salary-distributions)). Applying a log transformation makes the distribution more symmetric and closer to a normal shape (Figure \@ref(fig:salary-distributions)). Additionally, rookies are clearly at a disadvantage in terms of salary compared to more experienced players (Figure \@ref(fig:salary-distributions)).

```{r salary-distributions, echo=FALSE, fig.width=12, fig.height=6, fig.cap="Salary distributions and boxplots by player category"}

# 1) Helper to make a histogram
make_histogram <- function(data, var, title, xlab, fill_color) {
  ggplot(data, aes(x = {{ var }})) +
    geom_histogram(bins = 30, fill = fill_color, color = "white") +
    labs(title = title, x = xlab, y = "Count") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title  = element_text(face = "bold"),
      axis.title  = element_text(face = "bold")
    )
}

# 2) Helper to make a boxplot (by player_category)
make_boxplot <- function(data, var, title, ylab, fill_colors) {
  ggplot(data, aes(x = player_category, y = {{ var }}, fill = player_category)) +
    geom_boxplot() +
    scale_fill_manual(values = fill_colors) +
    labs(title = title, x = "Player Type", y = ylab) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position     = "none",
      plot.title          = element_text(face = "bold"),
      axis.title          = element_text(face = "bold")
    )
}

# 3) Use them
p1 <- make_histogram(nba, Salary/1e6,    "Salary Distribution",    "Salary (Millions USD)", "steelblue")
p2 <- make_histogram(nba, LogSalary,     "Log Salary Distribution","Log(Salary)",           "darkorange")
p3 <- make_histogram(nba, BoxCoxSalary,  "Box-Cox Salary Distribution", "Box-Cox(Salary)",  "springgreen")

colors <- c("skyblue","red","gold")
p4 <- make_boxplot(nba, Salary/1e6,   "Salary",     "Salary (Millions USD)", colors)
p5 <- make_boxplot(nba, LogSalary,    "Log Salary", "Log(Salary)",           colors)
p6 <- make_boxplot(nba, BoxCoxSalary, "Box-Cox Salary", "Box-Cox(Salary)",   colors)

# 4) Combine
(p1 + p2 + p3) / (p4 + p5 + p6)
```

## Performance Metric Distributions

PTS and FG are both right-skewed and highly correlated, while MP shows a fairly uniform distribution (Figure \@ref(fig:performance-dist)). True Shooting Percentage (TS%), calculated as TS% = PTS / (2 * (FGA + 0.44 * FTA))—which integrates scoring from field goals and free throws—tends to cluster around 0.6 regardless of salary (Figure \@ref(fig:performance-dist)). Notably, there's an outlier with a TS% of 0, and interestingly, an individual with a TS% of 0.8 isn’t earning much.

```{r performance-dist, echo=FALSE, fig.width=12, fig.height=4, fig.cap="Performance metric distributions"}

# Define function to create scatter plots with consistent aesthetics
create_scatter_plot <- function(data, x_var, y_var, fill_var, title, x_lab, y_lab) {
  ggplot(data, aes(x = {{x_var}}, y = {{y_var}})) +
    geom_point(aes(fill = {{fill_var}}), color = "black", alpha = 0.7, size = 3, shape = 21, stroke = 0.8) +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dotted") +
    scale_fill_manual(values = c("skyblue", "red", "gold"),
                      labels = c("Regular", "Rookie", "Star")) +
    labs(title = title, x = x_lab, y = y_lab, fill = "Player Type") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold"),
          axis.title = element_text(face = "bold"))
}

p1 <- make_histogram(nba,PTS,"Points Per Game (PTS)","PTS","forestgreen")
p2 <- make_histogram(nba,MP,"Minutes Played","MP","purple")
p3 <- make_histogram(nba,FG,"Field Goals","FG","brown")

p4 <- create_scatter_plot( nba, `TS%`, Salary/1e6, player_category,
                          "TS% vs Salary", "True Shooting %", "Salary (Millions)")

(p1 + p2 + p3 + p4) + plot_layout(nrow = 1)
```

## Pairwise Scatter Plots

Strong correlations among metrics—such as PTS, MP, WS, and FG—suggest these variables tend to move together (see Appendix Figure \@ref(fig:appendix-pairwise)). Additionally, histograms reveal significant right skewness, with TVMS displaying an almost bimodal distribution. Therefore, it's crucial to test for variance inflation factors (VIF) before constructing the models to ensure multicollinearity isn't an issue.

## Correlation Matrix

There is a noticeable positive correlation between points (PTS) and other performance metrics such as minutes played (MP), field goals (FG), turnovers (TOV), and win shares (WS) (see Appendix Figure \@ref(fig:appendix-corr)). This suggests that market-driven factors captured by TVMS are distinct from on-court performance.

## Salary vs. Key Predictors

The analysis indicates that performance metrics such as points per game, minutes played, and win shares are strongly correlated with salary, although there are exceptions—such as a rookie with high minutes but a relatively low salary (Appendix Figure \@ref(fig:appendix-salary-predictors)). Additionally, salaries tend to rise with years of experience before leveling off, highlighting the impact of tenure and potentially age. There are also clear positional differences; for example, centers and shooting guards tend to have lower median salaries compared to power forwards, who generally command higher median earnings. While there is an observation that players on larger market teams may earn more on average, this relationship appears to be complex and could benefit from further evaluation.

## Preliminary Models
To explore linear relationships between NBA player salaries and various performance metrics, we formulated nine models across three salary transformations. These models assess how different variables influence player compensation during the 2022-23 season.

### Model Specifications
We test three model specifications with three different dependent variables:

- **Models 1-3**: Raw Salary

- **Models 4-6**: Log-transformed Salary (LogSalary)

- **Models 7-9**: Box-Cox transformed Salary (BoxCoxSalary)

For each transformation, we test three specifications:

1. **Basic Model**: Only Points Per Game (PTS)

2. **Performance Model**: PTS, Total Rebounds (TRB), and Assists (AST)

3. **Comprehensive Model**: Performance metrics plus Experience (Exp_num) and TV Market Size (TVMS)

```{r prelim models, echo=FALSE}
model_1 <- lm(Salary ~ PTS, data = nba)
model_2 <- lm(Salary ~ PTS + TRB + AST, data = nba)
model_3 <- lm(Salary ~ PTS + TRB + AST + Exp_num + TVMS, data = nba)

model_4 <- lm(LogSalary ~ PTS, data = nba)
model_5 <- lm(LogSalary ~ PTS + TRB + AST, data = nba)
model_6 <- lm(LogSalary ~ PTS + TRB + AST + Exp_num + TVMS, data = nba)

model_7 <- lm(BoxCoxSalary ~ PTS, data = nba)
model_8 <- lm(BoxCoxSalary ~ PTS + TRB + AST, data = nba)
model_9 <- lm(BoxCoxSalary ~ PTS + TRB + AST + Exp_num + TVMS, data = nba)
```

## Collinearity Issues

No collinearity concern with the VIF results (<5), see Appendix Table \@ref(tab:appendix-vif).

## Heteroscedasticity & Robust Standard Error

From the fitted value v.s. resudul plot, whereas the first three models exhibit heteroscedasticity, the next six (using log and Box-Cox transformations) are homoscedastic.
These observations are further supported by the Breusch-Pagan test results.
(See Appendix Figure \@ref(fig:appendix-hetero) and Appendix Table \@ref(tab:appendix-bp).)

```{r robust std error, echo=FALSE}
robust_se_3 <- coeftest(model_3, vcov = vcovHC(model_3))
robust_se_6 <- coeftest(model_6, vcov = vcovHC(model_6))
robust_se_9 <- coeftest(model_9, vcov = vcovHC(model_9))
```

```{r define_stargazer_function, include=FALSE}
# Function to create stargazer tables with consistent formatting
create_stargazer_table <- function(models, title, column_labels, covariate_labels, table_label) {
  stargazer(models,
            type           = "html",
            title          = title,
            column.labels  = column_labels,
            covariate.labels = covariate_labels,
            model.numbers  = FALSE,
            align          = TRUE,
            font.size      = "small",
            no.space       = TRUE,
            label          = table_label,
            header         = FALSE)
}
```

## Model Selection Rationale

Based on the exploratory models (Appendix Tables \@ref(tab:appendix-models1), \@ref(tab:appendix-models2), and \@ref(tab:appendix-models3)), 
Model 3 (Raw Salary ~ PTS + TRB + AST + Exp_num + TVMS) with additional selected predictive variables provides improvements in R^2 compared to Model 1 and Model 2.
While log and Box-Cox transformations (Models 4 - 9) successfully decreased heteroscedasticity, they also decreased R^2 and make interpreting coefficients less direct (Appendix Figure \@ref(fig:appendix-hetero)). 
Model 3 includes key performance metrics, experience, and market size, all showing significance (especially with robust standard errors). Since we can use robust standard errors to address heteroscedasticity, we can still use the raw salary model (Model 3) for interpretability and significance testing.
(Appendix Table \@ref(tab:appendix-models-robust)). Therefore, we proceed with Model 3 as final model and apply confirmatory dataset to validate our findings.

# Confirmatory Analysis, Final Model, and Interpretation

This regression explains about 69% of the variation in NBA salaries, with scoring (PTS), experience (Expnum), and TV market size (TVMS) emerging as the most influential factors. Specifically, each additional point per game is associated with roughly a \$942 K boost in salary, each year in the league adds about \$1.01 M, and playing in a larger TV market contributes roughly \$642 K, all else held equal. Rebounds and assists do not carry significant weight once scoring is in the model, and visual diagnostics show that predictions align well with actual salaries for most players, though superstar contracts at the high end exhibit greater dispersion.

```{r final model, echo=FALSE}

# Load confirmatory dataset
confirm <- read_csv("../../data_nba/confirmatory.csv")
confirm <- confirm %>%
  filter(!is.na(Salary), GP > 0) %>%
  select(Salary, PTS, TRB, AST, Exp, tv_market_size) %>%
  mutate(
    LogSalary = log(Salary),
    BoxCoxSalary = bcPower(Salary, lambda_obj$lambda),
    Exp_num = as.numeric(ifelse(Exp == 'R', 0, Exp)),  # Create Exp_num column
    TVMS = tv_market_size,
    IsRookie = ifelse(Exp == 0, 1, 0),
  )

# Remove any remaining NAs to ensure consistent row counts
confirm <- na.omit(confirm)

# # Check for any remaining NAs
# na_summary_confirm <- colSums(is.na(confirm))
# if(sum(na_summary_confirm) > 0) {
#   cat("Variables with missing values in confirmatory dataset:\n")
#   print(na_summary_confirm[na_summary_confirm > 0])
# } else {
#   cat("No missing values remain in the confirmatory dataset.\n")
# }

# Fit the final model on confirmatory data
final_model <- lm(Salary ~ PTS + TRB + AST + Exp_num + TVMS, data = confirm)

# Calculate robust standard errors
robust_se_final <- coeftest(final_model, vcov = vcovHC(final_model))

# Plot predicted vs actual values
pred_vs_actual <- ggplot(data.frame(
  Actual = confirm$Salary/1e6,
  Predicted = predict(final_model)/1e6
), aes(x = Predicted, y = Actual)) +
  geom_point(color = "black", fill = "skyblue", alpha = 0.7, size = 3, shape = 21, stroke = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs Actual Salary",
       x = "Predicted Salary (Millions USD)",
       y = "Actual Salary (Millions USD)") +
  theme_minimal()

# Residual plot
resid_plot <- ggplot(data.frame(
  Fitted = fitted(final_model)/1e6,
  Residuals = residuals(final_model)/1e6
), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black", fill = "skyblue", alpha = 0.7, size = 3, shape = 21, stroke = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values (Millions USD)",
       y = "Residuals (Millions USD)") +
  theme_minimal()
```

# Model Limitations and Future Directions for Improvement

As potential next steps, our model could be refined by incorporating longitudinal data to capture temporal trends in player performance. Additionally, addressing omitted variable bias by including contextual factors—such as changes in league policy, evolving salary structures, the influence of agents, and injury history—may further enhance explanatory power. Stratified modeling, such as developing team-specific or region-specific models, could also improve fit and interpretability. Lastly, exploring nonlinear relationships through the inclusion of quadratic or interaction terms may better capture the complexity inherent in the data.

```{=latex}
\begin{figure}[!ht]
  \centering
  %————— LEFT BOX (table) —————
  \begin{minipage}[c]{0.48\textwidth}
    \centering
```
```{r table-side, results='asis', echo=FALSE}
stargazer(
  final_model,
  type           = "html",
  float          = FALSE,
  se             = list(robust_se_final[, "Std. Error"]),
  add.lines      = list(
    c("Robust p-val: PTS",    round(robust_se_final["PTS",    "Pr(>|t|)"], 3)),
    c("Robust p-val: TRB",    round(robust_se_final["TRB",    "Pr(>|t|)"], 3)),
    c("Robust p-val: AST",    round(robust_se_final["AST",    "Pr(>|t|)"], 3)),
    c("Robust p-val: Exp",    round(robust_se_final["Exp_num","Pr(>|t|)"], 3)),
    c("Robust p-val: TVMS",   round(robust_se_final["TVMS",   "Pr(>|t|)"], 3))
  ),
  title          = "Final Model on Confirmatory Dataset",
  column.labels  = "Final Model",
  covariate.labels = c("PTS","TRB","AST","Exp$_{\\text{num}}$","TVMS"),
  model.numbers  = FALSE,
  align          = TRUE,
  font.size      = "small",
  no.space       = TRUE,
  label          = "tab:final_model_side",
  header         = FALSE
)
```

```{=latex}
  \end{minipage}
  \hfill
  %——— RIGHT BOX (plots) ———
  \begin{minipage}[c]{0.48\textwidth}
    \centering
```
```{r plot-side, echo=FALSE, fig.width=4, fig.height=6}
# Combine your two ggplots
pred_vs_actual / resid_plot
```
```{=latex}
  \end{minipage}
  \caption{Final Model Results: (a) Coefficients with robust SE; (b) Predicted vs Actual and residual plots.}
  \label{fig:final_model_side}
\end{figure}
```
\FloatBarrier


# Appendix

## Pairwise Scatter-Plot Matrix {#fig:appendix-pairwise}
```{r appendix-pairwise-scatter, echo=FALSE, fig.width=12, fig.height=12, fig.cap="Pairwise scatter plot matrix of key NBA player metrics"}

# # Select key numerical variables for the pairwise plot - reducing to most important variables
# nba_subset_numeric <- nba %>%
#   select(
#     Salary, PTS, MP, FG, TRB, `3P%`, TOV, WS, AST, GP, TVMS, `TS%`, Exp_num, Age
#     ) # Reduced variable set for better readability

# # Create the enhanced pairwise scatter plot matrix
# GGally::ggpairs(
#   nba_subset_numeric,
#   upper = list(continuous = wrap("cor", size = 4, color = "darkblue")),
#   lower = list(continuous = wrap("points", alpha = 0.5, color = "#3366FF", size = 0.8)),
#   diag = list(continuous = wrap("densityDiag", fill = "lightblue", color = "darkblue")),
#   title = "Relationships Between Key NBA Player Metrics",
#   axisLabels = "show"
# ) +
#   theme_minimal(base_size = 12) +
#   theme(
#     axis.text = element_text(size = 10),
#     axis.title = element_text(size = 12, face = "bold"),
#     strip.text = element_text(size = 11, face = "bold"),
#     plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
#   )
```
\FloatBarrier

## Correlation Matrix {#fig:appendix-corr}

```{r corr-plot, echo=FALSE, fig.width=10, fig.height=8, fig.cap="Correlation matrix of key NBA player metrics"}
# nba %>%
#   select(
#     Salary, PTS, MP, FG, TRB, `3P%`, TOV, WS, AST, GP, TVMS, `TS%`, RookieOrStar, Exp_num, Age
#   ) %>%
#   GGally::ggcorr(label=TRUE, label_round=2, label_size=2.5)
```
\FloatBarrier


## Collinearity Diagnostics: VIF Comparison {#tab:appendix-vif}

```{r appendix-vif-table, echo=FALSE}
# Function to calculate VIF and format the output
calculate_vif <- function(model, model_name) {
  vif_df <- vif(model) %>%
    as.data.frame() %>%
    rownames_to_column("Variable") %>%
    rename(!!model_name := 2)
  return(vif_df)
}

# Calculate VIF for each model
vif_2 <- calculate_vif(model_2, "Model 2")
vif_3 <- calculate_vif(model_3, "Model 3")
vif_5 <- calculate_vif(model_5, "Model 5")
vif_6 <- calculate_vif(model_6, "Model 6")
vif_8 <- calculate_vif(model_8, "Model 8")
vif_9 <- calculate_vif(model_9, "Model 9")

# Join all VIF tables
vif_table <- list(vif_2, vif_3, vif_5, vif_6, vif_8, vif_9) %>%
  reduce(full_join, by = "Variable")

vif_table %>%
  kable(
    caption = "VIF Comparison Across Models (Excluding Models 1, 4, and 7)",
    booktabs = TRUE, digits = 2,
    col.names = c("Variable", "Model 2", "Model 3", "Model 5", "Model 6", "Model 8", "Model 9")
  ) %>%
  kable_styling(latex_options = "hold_position", position = "center")
```
\FloatBarrier

## Residual Diagnostics for Heteroscedasticity {#fig:appendix-hetero}

```{r appendix-hetero-plot, echo=FALSE, fig.pos='H', fig.cap="Residual plots for preliminary models", fig.width=12, fig.height=8}

create_resid_plot <- function(model, title) {
  ggplot(data.frame(fitted = fitted(model), resid = resid(model)), 
         aes(x = fitted, y = resid)) +
    geom_point(color = "black", fill = "skyblue", alpha = 0.7, size = 3, shape = 21, stroke = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = title, x = "Fitted Values", y = "Residuals") +
    theme_minimal(base_size = 8) +
    theme(
      plot.title = element_text(size = 8)  # ensure title font size is consistent
    )
}

# Create residual plots for all models
p1 <- create_resid_plot(model_1, "Model 1: Salary ~ PTS")
p2 <- create_resid_plot(model_2, "Model 2: Salary ~ PTS + TRB + AST")
p3 <- create_resid_plot(model_3, "Model 3: Salary ~ PTS + TRB + AST + Exp_num + TVMS")

p4 <- create_resid_plot(model_4, "Model 4: LogSalary ~ PTS")
p5 <- create_resid_plot(model_5, "Model 5: LogSalary ~ PTS + TRB + AST")
p6 <- create_resid_plot(model_6, "Model 6: LogSalary ~ PTS + TRB + AST + Exp_num + TVMS")

p7 <- create_resid_plot(model_7, "Model 7: BoxCoxSalary ~ PTS")
p8 <- create_resid_plot(model_8, "Model 8: BoxCoxSalary ~ PTS + TRB + AST")
p9 <- create_resid_plot(model_9, "Model 9: BoxCoxSalary ~ PTS + TRB + AST + Exp_num + TVMS")

# Display plots in a grid
(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```
\FloatBarrier

## Breusch–Pagan Test Results {#tab:appendix-bp}

```{r bp-test, echo=FALSE}
# Perform Breusch-Pagan test for heteroscedasticity
bp_results <- data.frame(
  Model = paste("Model", 1:9),
  BP_Statistic = numeric(9),
  P_Value = numeric(9),
  Conclusion = character(9)
)

# Function to run BP test and store results
run_bp_test <- function(model, index) {
  bp_test <- bptest(model)
  bp_results$BP_Statistic[index] <<- round(bp_test$statistic, 3)
  bp_results$P_Value[index] <<- round(bp_test$p.value, 4)
  bp_results$Conclusion[index] <<- ifelse(bp_test$p.value < 0.05,
                                         "Heteroscedastic",
                                         "Homoscedastic")
}

# List of models to test
models <- list(model_1, model_2, model_3, model_4, model_5, model_6, model_7, model_8, model_9)

# Run tests for all models using a loop
for (i in 1:length(models)) {
  run_bp_test(models[[i]], i)
}

# Display BP test results
bp_results %>%
  kable(caption = "Breusch-Pagan Test Results for Heteroscedasticity",
        booktabs = TRUE,
        col.names = c("Model", "BP Statistic", "P-Value", "Conclusion")) %>%
  kable_styling(latex_options = "hold_position", position = "center")
```
\FloatBarrier


## Salary vs. Key Predictors {#fig:appendix-salary-predictors}

```{r salary-predictors, echo=FALSE, fig.pos='H', fig.cap="Salary vs key predictors", fig.width=12, fig.height=12}
# The player_category is already defined above, so we don't need to redefine it
# Add this code before the plot to check
# cat("Number of All-Star players in the dataset:", sum(nba$star_player == 1), "\n")

# Create scatter plots using the function
p_pts <- create_scatter_plot(nba, PTS, Salary/1e6, player_category,
                             "Salary vs. Points Per Game", "Points Per Game", "Salary (Millions USD)")
p_mp <- create_scatter_plot(nba, MP, Salary/1e6, player_category,
                            "Salary vs. Minutes Played", "Minutes Played", "Salary (Millions USD)")
p_ws <- create_scatter_plot(nba, WS, Salary/1e6, player_category,
                            "Salary vs. Win Shares", "Win Shares", "Salary (Millions USD)")

# 1) Factor Team in one step, sorted alphabetically
nba <- nba %>%
  mutate(Team = factor(Team, levels = sort(unique(Team))))

# Bar plot for TV Market Size remains the same
p_tvms <- ggplot(nba, aes(x= Team, y=TVMS)) +
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="TV Market Size by Team", x="Team", y="TV Market Size") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(face="bold"),
        axis.title = element_text(face="bold"))

# Box plot for Salary by Team remains the same
p_team <- ggplot(nba, aes(x= Team, y=Salary/1e6)) +
  geom_boxplot(fill="lightgreen") +
  labs(title="Salary by Team", x="Team", y="Salary (Millions USD)") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size=10),
        plot.title = element_text(face="bold"),
        axis.title = element_text(face="bold"))

# Box plot for Salary by Position remains the same
p_pos <- ggplot(nba, aes(x = Position, y = Salary/1e6, fill = Position)) +
  geom_boxplot() +
  labs(title = "Salary by Position", x = "Position", y = "Salary (Millions USD)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face="bold"),
        axis.title = element_text(face="bold"))

# Use create_scatter_plot function for Salary vs Experience plot
p_exp <- create_scatter_plot(nba, Exp_num, Salary/1e6, player_category,
                            "Salary vs. Experience", "Years of Experience", "Salary (Millions USD)")

# Display the plots in a grid layout with the team plots stacked vertically
(p_pts + p_mp + p_ws) / (p_tvms + p_pos) / (p_team + p_exp)
```
\FloatBarrier
\newpage

## Raw‐Salary Exploratory Models {#tab:appendix-models1}

```{r model123 in stargazer, results='asis', echo=FALSE, message = FALSE, warning = FALSE}
create_stargazer_table(list(model_1, model_2, model_3),
                       "NBA Salary preliminary Models (Raw Salary)",
                       c("Model 1", "Model 2", "Model 3"),
                       c("PTS", "TRB", "AST", "Exp$_{\\text{num}}$", "TVMS"),
                       "tab:models1")
```
\FloatBarrier
\newpage

## Log‐Salary Exploratory Models {#tab:appendix-models2}

```{r model456 in stargazer, results='asis', echo=FALSE, message = FALSE, warning = FALSE}
create_stargazer_table(list(model_4, model_5, model_6),
                       "NBA Salary preliminary Models (Log Salary)",
                       c("Model 4", "Model 5", "Model 6"),
                       c("PTS", "TRB", "AST", "Exp$_{\\text{num}}$", "TVMS"),
                       "tab:models2")
```
\FloatBarrier
\newpage

## Box-Cox-Salary Exploratory Models {#tab:appendix-models3}

```{r model789 in stargazer, results='asis', echo=FALSE, message = FALSE, warning = FALSE}
create_stargazer_table(list(model_7, model_8, model_9),
                       "NBA Salary preliminary Models (Box-Cox Salary)",
                       c("Model 7", "Model 8", "Model 9"),
                       c("PTS", "TRB", "AST", "Exp$_{\\text{num}}$", "TVMS"),
                       "tab:models3")
```
\FloatBarrier
\newpage

## Robust SE for Selected Models {#tab:appendix-models-robust}

```{r robust model369 in stargazer with robust se, results='asis', echo=FALSE, message = FALSE, warning = FALSE}

pvals_3 <- robust_se_3[, "Pr(>|t|)"]
pvals_6 <- robust_se_6[, "Pr(>|t|)"]
pvals_9 <- robust_se_9[, "Pr(>|t|)"]

# Display model summary with robust standard errors
stargazer(model_3, model_6, model_9,
          type = "html",
          se = list(robust_se_3[, "Std. Error"],
                    robust_se_6[, "Std. Error"],
                    robust_se_9[, "Std. Error"]),
          add.lines = list(
            c("Robust p-val: PTS",
              round(pvals_3["PTS"], 3), round(pvals_6["PTS"], 3), round(pvals_9["PTS"], 3)),
            c("Robust p-val: TRB",
              round(pvals_3["TRB"], 3), round(pvals_6["TRB"], 3), round(pvals_9["TRB"], 3)),
            c("Robust p-val: AST",
              round(pvals_3["AST"], 3), round(pvals_6["AST"], 3), round(pvals_9["AST"], 3)),
            c("Robust p-val: Exp$_{\\text{num}}$",
              round(pvals_3["Exp_num"], 3), round(pvals_6["Exp_num"], 3), round(pvals_9["Exp_num"], 3)),
            c("Robust p-val: TVMS",
              round(pvals_3["TVMS"], 3), round(pvals_6["TVMS"], 3), round(pvals_9["TVMS"], 3))
          ),
          title = "NBA Salary Models with Robust Standard Errors",
          column.labels = c("Model 3", "Model 6", "Model 9"),
          covariate.labels = c("PTS", "TRB", "AST", "Exp$_{\\text{num}}$", "TVMS"),
          model.numbers = FALSE,
          align = TRUE,
          font.size = "small",
          no.space = TRUE,
          label = "tab:models_robust",
          header = FALSE)

```
\FloatBarrier
\newpage

## Data Dictionary

| Variable | Description |
|----------|-------------|
| Salary   | Player salary in USD |
| LogSalary | Natural logarithm of salary |
| PTS      | Points per game |
| MP       | Minutes played per game |
| FG       | Field goals made per game |
| TRB      | Total rebounds per game |
| `3P%`    | 3-point shooting percentage |
| TOV      | Turnovers per game |
| WS       | Win shares |
| GP       | Games played |
| AST      | Assists per game |
| TVMS     | TV Market Size |
| TS%      | True Shooting Percentage |
| star_player | All-Star status (1=Yes, 0=No) |
| Position | Player position |
| Age      | Player age |
| Exp_num  | Years of NBA experience |


# References

1. Southwest Journal. (n.d.). *NBA Players’ Salaries: How Much Do NBA Players Make?* Retrieved from [https://www.southwestjournal.com/sport/nba/nba-players-salaries/?utm_source=chatgpt.com](https://www.southwestjournal.com/sport/nba/nba-players-salaries/?utm_source=chatgpt.com)

2. Papadaki, I., & Tsagris, M. (2020). *Are NBA Players’ Salaries in Accordance with Their Performance on Court?* In *Advances in Econometrics, Operational Research, Data Science and Actuarial Studies* (pp. 405–428). [https://doi.org/10.1007/978-3-030-85254-2_25](https://doi.org/10.1007/978-3-030-85254-2_25)

3. Sigler, K.J., & Sackley, W.H. (2000). *NBA Players: Are They Paid for Performance?* *Managerial Finance, 26*(7), 46–51. [https://doi.org/10.1108/03074350010766783](https://doi.org/10.1108/03074350010766783)